Contextuality and Nonlocality in Discrete Quantum Theory

### 2016-02-09 DQC3.tex
"6.1 Gleason's Theorem in DQT" S-7 Parallelism

Conversely, any ${\cal P}$ with these two properties can be identified
with a mixed state $\rho$ by Gleason's theorem~\cite{Redhead1987-REDINA,gleason1957,peres1995quantum}\footnote{In order to simplify the statement, we consider the probability on
the state vectors instead of the projectors or closed subspaces as
in the Gleason's original paper.}.
\begin{thm}[Gleason's]
For $d\ge3$, if ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$ is
invariant under scalar multiplication and is a probability, i.e.,
\begin{eqnarray}
 &  & {\cal P}(\ket{i})={\cal P}(\gamma\ket{i})\textrm{ for any vector }\ket{i}\in\C^{d*}\textrm{, and }\gamma\in\C^{*}\textrm{ ,}\label{eq:parallel-DQT(I)-1}\\
 &  & {\displaystyle \sum_{i=0}^{d-1}{\cal P}\left(\ket{i}\right)=1}\textrm{ for any orthonormal basis }\{\ket{i}\}\textrm{ ,}\label{eq:total-probability-1}
\end{eqnarray}
there is a density matrix $\rho$ such that ${\cal P}={\cal P}_{\rho}$.
\end{thm}
Gleason's theorem requires $d\ge3$ because different orthonormal
bases may share their vectors. For example, $\left\{ \ket{0},\ket{1},\ket{2}\right\} $
and $\left\{ \ket{0},\frac{1}{\sqrt{2}}\left(\ket{1}+\ket{2}\right),\frac{1}{\sqrt{2}}\left(\ket{1}-\ket{2}\right)\right\} $
share the vector $\ket{0}$. In contrast, when $d=2$, two orthonormal
bases are either identical or completely different after we use equation
(\ref{eq:parallel-DQT(I)-1}) to identify ${\cal P}\left(\ket{i}\right)$
and ${\cal P}(\gamma\ket{i})$. Therefore, when we solve the system
of linear equation (\ref{eq:total-probability-1}), we actually solve
them individually and may get a discontinuous solution. For example,
consider 
\[
{\cal P}
\left(\begin{array}{c} \cos\theta\\ \sin\theta\rme^{\rmi\gamma} \end{array}\right)=
\cases{ 
0 & , if $0 \le \theta < \frac{\pi}{2}$\\
1 & , if $\frac{\pi}{2} \le \theta < \pi$ } 
\]
${\cal P}$ satisfies equation (\ref{eq:parallel-DQT(I)-1}) and (\ref{eq:total-probability-1})
because we have 
\begin{eqnarray*}
 &  & {\cal P}\left(\begin{array}{c}
\cos\theta\\
\sin\theta\rme^{\rmi\gamma}
\end{array}\right)+{\cal P}\left(\begin{array}{c}
-\sin\theta\rme^{-\rmi\gamma}\\
\cos\theta
\end{array}\right)\\
 & = & {\cal P}\left(\begin{array}{c}
\cos\theta\\
\sin\theta\rme^{\rmi\gamma}
\end{array}\right)+{\cal P}\left(\begin{array}{c}
\cos\left(\theta+\frac{\pi}{2}\right)\\
\sin\left(\theta+\frac{\pi}{2}\right)\rme^{\rmi\gamma}
\end{array}\right)=1\textrm{ .}
\end{eqnarray*}
However, ${\cal P}_{\rho}$ is always continuous. Therefore, no density
matrix $\rho$ satisfies ${\cal P}={\cal P}_{\rho}$.

Although the situation is different between $d=2$ and $d\ge3$ in
CQT, a probability in DQT cannot be identified with a state despite
the dimension of the system. In order to clearly state the theorem,
the discrete Born rule $\dpr\left(\ket{\Psi},\ket{i}\right)$ will
be written in the curried form $\dpr_{\Psi}\left(\ket{i}\right)$.
Then, we have 
\begin{thm}
\label{thm:GleasonDQT}For any $d\ge2$, there exists $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}_{2}$
which is invariant under scalar multiplication and is a probability,
i.e., 
\begin{eqnarray}
 &  & \hat{\pi}(\ket{i})=\hat{\pi}(\gamma\ket{i})\textrm{ for any vector }\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{ ,}\label{eq:parallel-DQT}\\
 &  & {\displaystyle \bigvee_{i=0}^{d-1}\hat{\pi}\left(\ket{i}\right)=}\textrm{possible for any orthonormal basis }\{\ket{i}\}\textrm{ , }\label{eq:total-probability-DQT}
\end{eqnarray}
such that $\dpr_{\Psi}\ne\hat{\pi}$ for any $\ket{\Psi}\in\ffzd{p^{2}}$.\end{thm}
\begin{proof}
Fix a basis $\{\ket{i}\}_{i=0}^{d-1}$. Consider 
\[
{\hat{\pi}}\left(\ket{\Phi}\right)=
\cases{ 
\textrm{impossible} & , if $\ip{\Phi}{0} = 0$ or $\ip{\Phi}{0} = \ldots = \ip{\Phi}{d-1}\ne 0$ ; \\
\textrm{possible} & , otherwise. } 
\]
$\hat{\pi}$ satisfies equation (\ref{eq:parallel-DQT}) because we
have 
\begin{eqnarray*}
 &  & \hat{\pi}\left(\ket{\Phi}\right)=\textrm{impossible}\\
 & \Leftrightarrow & \ip{\Phi}{0}=0\textrm{ or }\ip{\Phi}{0}=\ldots=\ip{\Phi}{d-1}\ne0\\
 & \Leftrightarrow & \gamma\ip{\Phi}{0}=0\textrm{ or }\gamma\ip{\Phi}{0}=\ldots=\gamma\ip{\Phi}{d-1}\ne0\\
 & \Leftrightarrow & \hat{\pi}\left(\gamma\ket{\Phi}\right)=\textrm{impossible}
\end{eqnarray*}
for any $\gamma\in\ff{p^{2}}^{*}$. For equation (\ref{eq:total-probability-DQT}),
because $\textrm{impossible}\lor\textrm{possible}=\textrm{possible}$,
it is sufficient to prove for any $\ket{\Phi}\in\ffzd{p^{2}}$ such
that $\hat{\pi}\left(\ket{\Phi}\right)=\textrm{impossible}$, there
is a $\ket{\Psi}\in\ffzd{p^{2}}$ such that $\ip{\Phi}{\Psi}=0$ and
$\hat{\pi}\left(\ket{\Psi}\right)=\textrm{possible}$.
\begin{casenv}
\item When $\ip{\Phi}{0}=0$, $\ket{0}$ is what we want because $\ip{0}{0}=1\ne0$
implies $\hat{\pi}\left(\ket{0}\right)=\textrm{possible}$.
\item When $\ip{\Phi}{0}=\ldots=\ip{\Phi}{d-1}\ne0$, let $\alpha=\ip{\Phi}{0}$,
and consider the state 
\[
\ket{\Psi}=\left(\begin{array}{ccccc}
\alpha & -\alpha & 0 & \ldots & 0\end{array}\right)^{T}\textrm{ .}
\]
Then, $\ip{\Phi}{\Psi}=\alpha^{*}\alpha-\alpha^{*}\alpha=0$. Moreover,
$\ket{\Psi}{0}=\alpha^{*}\ne0$ and $\ket{\Psi}{0}=\alpha^{*}\ne-\alpha^{*}=\ket{\Psi}{1}$
implies $\hat{\pi}\left(\ket{\Psi}\right)=\textrm{possible}$.
\end{casenv}
Therefore, equation (\ref{eq:total-probability-DQT}) holds.

>>>>

Conversely, any ${\cal P}$ with these two properties can be identified
with a mixed state $\rho$ by Gleason's theorem~\cite{Redhead1987-REDINA,gleason1957,peres1995quantum}\footnote{In order to simplify the statement, we consider the probability on
the state vectors instead of the projectors or closed subspaces as
in the Gleason's original paper.}.
\begin{thm}[Gleason's]
For $d\ge3$, if ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$ is
invariant under scalar multiplication and is a probability, i.e.,
\begin{eqnarray}
 &  & {\cal P}(\ket{i})={\cal P}(\gamma\ket{i})\textrm{ for any vector }\ket{i}\in\C^{d*}\textrm{, and }\gamma\in\C^{*}=\C\backslash\left\{ 0\right\} \textrm{ ,}\label{eq:parallel-DQT(I)-1}\\
 &  & {\displaystyle \sum_{i=0}^{d-1}{\cal P}\left(\ket{i}\right)=1}\textrm{ for any orthonormal basis }\{\ket{i}\}\textrm{ ,}\label{eq:total-probability-1}
\end{eqnarray}
there is a density matrix $\rho$ such that ${\cal P}={\cal P}_{\rho}$.
\end{thm}
Gleason's theorem requires $d\ge3$ because different orthonormal
bases may share their vectors. For example, $\left\{ \ket{0},\ket{1},\ket{2}\right\} $
and $\left\{ \ket{0},\frac{1}{\sqrt{2}}\left(\ket{1}+\ket{2}\right),\frac{1}{\sqrt{2}}\left(\ket{1}-\ket{2}\right)\right\} $
share the vector $\ket{0}$. In contrast, when $d=2$, two orthonormal
bases are either identical or completely different after we use equation
(\ref{eq:parallel-DQT(I)-1}) to identify ${\cal P}\left(\ket{i}\right)$
and ${\cal P}(\gamma\ket{i})$. Therefore, when we solve the system
of linear equation (\ref{eq:total-probability-1}), we actually solve
them individually and may get a discontinuous solution. For example,
consider 
\[
{\cal P}
\left(\begin{array}{c} \cos\theta\\ \sin\theta\rme^{\rmi\gamma} \end{array}\right)=
\cases{ 
0 & , if $0 \le \theta < \frac{\pi}{2}$\\
1 & , if $\frac{\pi}{2} \le \theta < \pi$ } 
\]
${\cal P}$ satisfies equation (\ref{eq:parallel-DQT(I)-1}) and (\ref{eq:total-probability-1})
because we have 
\begin{eqnarray*}
 &  & {\cal P}\left(\begin{array}{c}
\cos\theta\\
\sin\theta\rme^{\rmi\gamma}
\end{array}\right)+{\cal P}\left(\begin{array}{c}
-\sin\theta\rme^{-\rmi\gamma}\\
\cos\theta
\end{array}\right)\\
 & = & {\cal P}\left(\begin{array}{c}
\cos\theta\\
\sin\theta\rme^{\rmi\gamma}
\end{array}\right)+{\cal P}\left(\begin{array}{c}
\cos\left(\theta+\frac{\pi}{2}\right)\\
\sin\left(\theta+\frac{\pi}{2}\right)\rme^{\rmi\gamma}
\end{array}\right)=1\textrm{ .}
\end{eqnarray*}
However, ${\cal P}_{\rho}$ is always continuous. Therefore, no density
matrix $\rho$ satisfies ${\cal P}={\cal P}_{\rho}$.

Although the situation is different between $d=2$ and $d\ge3$ in
CQT, a probability in DQT cannot be identified with a state despite
the dimension of the system. 
\begin{thm}
\label{thm:GleasonDQT}For any $d\ge2$, there exists $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}_{2}$
which is invariant under scalar multiplication and is a probability,
i.e., 
\begin{eqnarray}
 &  & \hat{\pi}(\ket{i})=\hat{\pi}(\gamma\ket{i})\textrm{ for any vector }\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{ ,}\label{eq:parallel-DQT}\\
 &  & {\displaystyle \bigvee_{i=0}^{d-1}\hat{\pi}\left(\ket{i}\right)=}\textrm{possible for any orthonormal basis }\{\ket{i}\}\textrm{ , }\label{eq:total-probability-DQT}
\end{eqnarray}
such that $\dpr_{\Psi}\ne\hat{\pi}$ for any $\ket{\Psi}\in\ffzd{p^{2}}$,
where $\dpr_{\Psi}\left(\ket{i}\right)=\dpr\left(\ket{\Psi},\ket{i}\right)$
is the discrete Born rule (\ref{eq:modality}).\end{thm}
\begin{proof}
Fix a basis $\{\ket{i}\}_{i=0}^{d-1}$. Consider 
\[
{\hat{\pi}}\left(\ket{\Phi}\right)=
\cases{ 
\textrm{impossible} & , if $\ip{\Phi}{0} = 0$ or $\ip{\Phi}{0} = \ldots = \ip{\Phi}{d-1}\ne 0$ ; \\
\textrm{possible} & , otherwise. } 
\]
$\hat{\pi}$ satisfies equation (\ref{eq:parallel-DQT}) because we
have 
\begin{eqnarray*}
 &  & \hat{\pi}\left(\ket{\Phi}\right)=\textrm{impossible}\\
 & \Leftrightarrow & \ip{\Phi}{0}=0\textrm{ or }\ip{\Phi}{0}=\ldots=\ip{\Phi}{d-1}\ne0\\
 & \Leftrightarrow & \gamma\ip{\Phi}{0}=0\textrm{ or }\gamma\ip{\Phi}{0}=\ldots=\gamma\ip{\Phi}{d-1}\ne0\\
 & \Leftrightarrow & \hat{\pi}\left(\gamma\ket{\Phi}\right)=\textrm{impossible}
\end{eqnarray*}
for any $\gamma\in\ff{p^{2}}^{*}$. For equation (\ref{eq:total-probability-DQT}),
because $\textrm{impossible}\lor\textrm{possible}=\textrm{possible}$,
it is sufficient to prove the following claim.
\begin{description}
\item [{Claim}] For any $\ket{\Phi}\in\ffzd{p^{2}}$ such that $\hat{\pi}\left(\ket{\Phi}\right)=\textrm{impossible}$,
there is a $\ket{\Psi}\in\ffzd{p^{2}}$ such that $\ip{\Phi}{\Psi}=0$
and $\hat{\pi}\left(\ket{\Psi}\right)=\textrm{possible}$.\end{description}
\begin{casenv}
\item When $\ip{\Phi}{0}=0$, $\ket{0}$ is what we want because $\ip{0}{0}=1\ne0$
and $\ip{0}{d-1}=0$ implies $\hat{\pi}\left(\ket{0}\right)=\textrm{possible}$.
\item When $\ip{\Phi}{0}=\ldots=\ip{\Phi}{d-1}\ne0$, let $\alpha=\ip{\Phi}{0}$,
and consider the state 
\[
\ket{\Psi}=\left(\begin{array}{ccccc}
\alpha & -\alpha & 0 & \ldots & 0\end{array}\right)^{T}\textrm{ .}
\]
Then, $\ip{\Phi}{\Psi}=\alpha^{*}\alpha-\alpha^{*}\alpha=0$. Moreover,
$\ket{\Psi}{0}=\alpha^{*}\ne0$ and $\ket{\Psi}{0}=\alpha^{*}\ne-\alpha^{*}=\ket{\Psi}{1}$
implies $\hat{\pi}\left(\ket{\Psi}\right)=\textrm{possible}$.
\end{casenv}
Therefore, equation (\ref{eq:total-probability-DQT}) holds.
